---
title: "Caracal Distribution"
author: "Grethen de Waal"
date: "`r Sys.Date()`"
output: html_document
---


The urban caracal experiences significant anthropogenic challenges that impact their territories, foraging and lifespans. Caracals living in the Cape Peninsula frequently encounter humans as they forage along the urban edge. 

<br/>

Distribution data was taken from two sources, iNaturalist and Movebank. 

<br/>

### iNaturalist sightings
<br/>

Read data into R using rinat package
```{r}
library("rinat")

inat_obs_raw <- get_inat_obs(taxon_name = "Caracal caracal",
                   bounds = c(-35, 18, -33.5, 18.5), 
                   quality = "research",
                   maxresults = 500)
```

<br/>

Some users obscured the location data of their sightings, so a number of observations show caracals in the middle of the ocean. For this reason, all observations with column "coordinates_obscured" being "false" were removed from the dataset.

```{r}
library(tidyverse)

inat_obs <- inat_obs_raw %>% 
  filter(coordinates_obscured == "false")
```

<br/>


The data frame was converted to spatial data:

```{r}
library(sf)

inat_sf <- st_as_sf(
  inat_obs, 
  coords = c("longitude", "latitude"), 
  crs = 4326)
```

<br/>

The data were plotted onto an interactive map:

```{r}
library(leaflet)
library(mapview)
library(leafpop)
library(move2)

sighting <- inat_sf %>% 
  mutate('🔗' = paste("<b><a href='", url, "'>Link to iNat observation</a></b>"),
         '📷' = paste("<img src ='", image_url , "' height='128'>")) %>% 
  rename('👤' = "user_login", captive = captive_cultivated)

mapview(sighting,
        popup = popupTable(sighting, 
                           zcol = c("👤", "🔗", "📷")))
```

**Figure 1** shows caracal sightings reported on [iNaturalist](https://www.inaturalist.org/observations?place_id=123155&subview=map&taxon_id=42042) in the Cape Peninsula. The map is interactive. Hovering over an observation ("sighting") displays a number. Clicking on the number displays the user who submitted the sighting, the URL to the iNaturalist entry and the photo that goes with it. 

<br/>


### Movebank data
The [Movebank database](https://www.movebank.org/cms/webapp?gwt_fragment=page=studies,path=study1832666571) where [Urban Caracal Project](https://www.urbancaracal.org/) (UCP) stores its distribution data has over 57,000 observations, so a single cat was selected to map. 

<br/>

Read data into R
```{r}
ucp_csv <- read.csv("Caracal movement ecology study in Cape Town, South Africa.csv")

ucp_sf <- ucp_csv %>%
  filter(!is.na(location.long)) %>%                             # Remove missing GPS points or it won't run
  filter(individual.local.identifier == "Jasper - TMC08") %>%   
  st_as_sf(coords = c("location.long", "location.lat"), 
           crs = 4326)

```

<br/>

Plot the Movebank data

```{r}
ucp_map <- leaflet() %>% 
  addTiles(group = "Default") %>% 
  addCircleMarkers(data = ucp_sf, 
                   radius = 0.5, 
                   color = "blue")

ucp_map
```

**Figure 2** shows the sightings of one of the tagged UCP cats called Jasper. This map is interactive with a zoom function.

<br/>

### Combined map

```{r}
library(ggspatial)

caracal_map <- ggplot() + 
  annotation_map_tile(type = "osm", progress = "none") + 
  geom_sf(data = ucp_sf, color = "orange") +
  geom_sf(data = inat_sf, color = "purple")

caracal_map
```

**Figure 3** shows Movebank's data on Jasper's movements (orange) and the iNaturalist sightings (purple) on a static map of the Cape Peninsula.